---
title: "WQD7004 Diamond Price Prediction and Clarity Classification Project"
author: "Yousef, put your names here"
date: '2022-06-13'
output: html_document
---

<!-- STUDENT 1 -->
<center> <h1><b>1.Introduction</b></h1> </center>

Diamond is a solid form of carbon element that present in crystal structure that known as diamond cubic making it unique. Diamond is known with their hardness, good thermal conductivity, high index of refraction, high dispersion, and adamantine luster. The high luster gives diamond the ability to reflect lights that strikes on their surface thus giving them the ‘sparkle’. 

Colour and clarity determine the price of diamond to be selected as jewelry gems.  Jewelry diamonds have the lowest number of specific gravity with it happens to be very close to 3.52 with minimal impurities and defects.
Quality of diamonds that are made into jewelry gems are determined by color, cut, clarity and carat weight. Diamond attributes are as follows: 

•	**Colour**: Most quality diamond ranging from colorless to slightly yellow, brown or grey. The highest and most valuable diamonds is the one that are completely colorless. 

•	**Clarity**: An ideal diamond is free from fracture and particles of foreign material within the gems as low clarity gems tends to degrade the appearance, reduce the strength of the stone thus lower its value.

•	**Cut**: Quality of designs and craftsmanship determines the appearance of diamonds that later determines the price. Angles of facets cut, proportions of design and quality of polishing determines face-up appearance, brilliance, scintillation, pattern and fire. A perfect diamond stones are perfectly polished, highly reflective, emit maximum amount of fire, faceted faces equal in size and the edges meet perfectly also identical in shape. 

•	**Carat**: A unit of weight equal to 1/5 of a gram or 1/142 of an ounce. Small diamonds are usually cost less per carat because of its common presences. 

Another category of diamonds that are currently becoming a trend among diamond jewelry lovers are colored diamonds that occur in variety of hues such as red, pink, yellow, orange, purple, blue, green, and brown. The quality of this diamond’s type is determined by intensity, purity, and quality of their colour, which, the most saturated and vivid colour hold a greater price.


<!-- STUDENT 1 -->
<center> <h1><b>2.Dataset</b></h1> </center>

•	**Title**: Diamonds 

•	**Year**: 2017

•	**Purpose of Dataset**: A great simple dataset for beginners who is learning to work in data analysis and visualization. 

•	**Content**: Diamond attributes of price, carat, cut, color, clarity, length, width, depth, total depth percentage, width of top of diamonds.

•	**Structure**: Mainly consist of integers, floating point values also string.

•	**Summary**: This dataset describes attributes of the 54,000 diamonds together with the price so the dataset can be make used to propose suitable linear regression or just normal exploratory data analysis. 


<!-- STUDENT 1 -->
<center> <h1><b>3.Objectives</b></h1> </center>

Diamond gems is one of the most popular gems in entire world. This valuable gem can be worth from as low as hundreds and up to millions. However, no clear understanding of why and how diamond gems can be so expensive. Therefore, exploring which attributes determine the value of a diamond gems may helps with predicting the price of the diamonds. Objectives of this project are stated as below: 

* 1. To explore which attributes contribute to the price range in diamond gems. 

* 2. To predict the price of diamond gems from corresponding attributes. 

<center> <h1><b>4.Data Exploration and Preprocessing</b></h1> </center>

### 4.1 First look at the data 



```{r}
# required packages

data = read.csv('data/dirty_dataset_diamond.csv')
summary(data)
```

From the above code, it's observed that there is an **unnamed attribute**, exploring it further

```{r}
data$Unnamed..0[1:20]
```


From the above output we find that it's simply an index to all the diamond observations we have in the data set, thus it can be dropped since it useless for our analysis.

```{r}
# dropping Unnamed..0 attribute
drops <- c('Unnamed..0')
data = data[ , !(names(data) %in% drops)]
```

### 4.2 Categorical Features Exploration

> In our dataset we have 3 categorical features **cut**, **color**, **clarity**, bar charts will be used to explore them

#### **cut** attribute exploration
```{r}
library(ggplot2)
ggplot(data, aes(x = factor(cut))) +
    geom_bar()
```

From the above figure, we can observer that most of the diamond cuts are either *Ideal* or *Very Good*, however it's also seen that there are *Unknown* cuts, this might correspond to missing data. Since the *Unknown* cuts represent only a small fraction of the all cuts, we will consider dropping all observations with that value, this will ensure that our Machine Learning models are more robust.
```{r}
data <- subset(data, cut != 'Unknown')
ggplot(data, aes(x = factor(cut))) +
    geom_bar()
```

Exploring if there are any correlation between the *cut* and the *price* of a diamond
```{r}
library(scales)
ggplot(data,
       aes(y = factor(cut,
                      labels = c("Fair",
                                 "Good",
                                 "Ideal",
                                 "Premium",
                                 "Very Good")), 
           x = price, 
           color = cut)) +
  geom_jitter(alpha = 0.7,
              size = 1.5) + 
  scale_x_continuous(label = dollar) +
  labs(title = "Price of Diamond by Cut", 
       x = "",
       y = "") +
  theme_minimal() +
  theme(legend.position = "none")
```

From the above figure we can see that diamonds with **Premium** and **Ideal** cuts are laying the end of the price spectrum

#### **color** attribute exploration
```{r}
ggplot(data, aes(x = factor(color))) +
    geom_bar()
```

From the above figure, we can see that there is almost a normal distribution, and there are no any missing values, moreover, exploring using boxplot we can see if there are any relationships between the **color** attribute and the target variable **price** 

#### **clarity** attribute exploration
```{r}
ggplot(data, aes(x = factor(clarity))) +
    geom_bar()
```

From the above figure, we can see that there are no any abnormalities or missing values, moreover to explore the relationship between **clarity** and the diamond **price** we will use boxplot
```{r}
library(dplyr)
plotdata <- data %>%
  group_by(clarity) %>%
  summarize(n = n(),
         mean = mean(price),
         sd = sd(price),
         se = sd / sqrt(n),
         ci = qt(0.975, df = n - 1) * sd / sqrt(n))

ggplot(plotdata, 
       aes(x = clarity, 
           y = mean, 
           group = 1)) +
  geom_point(size = 3) +
  geom_line() +
  geom_errorbar(aes(ymin = mean - se, 
                    ymax = mean + se), 
                width = .1)
```


From the above plot we can see that the average price of diamonds with **Sl2** cuts are the highest while the average price of **VVS1** cuts is the lowest

### 4.2 Interval Features Exploration
```{r}
summary(data)
```
From the above summary we can see that min value of "x", "y", "z" are zero this indicates that there are faulty values in data that represents dimensionless or 2-dimensional diamonds. So we need to filter out those as it clearly faulty data points. 

```{r}
library(tidyr)
data <- filter(data, x > 0, y > 0, z > 0)
summary(data)
```
From the above summary we can see that all 0's were successfully removed from the x, y and z attributes, and our total data points reduced from 53940 to 53855 which is not that much.

To explore the distribution of interval variables, histogram plots are used

```{r}
library("tidyr")
library("ggplot2")
drops <- c('cut', 'clarity', 'color')
hist_data = data[ , !(names(data) %in% drops)]
data_long <- hist_data %>%                          # Apply pivot_longer function
  pivot_longer(colnames(hist_data)) %>% 
  as.data.frame()
head(data_long) 
ggp1 <- ggplot(data_long, aes(x = value)) +    # Draw each column as histogram
  geom_histogram() + 
  facet_wrap(~ name, scales = "free")
ggp1
```

From the above histogram plots, we can see that the **price** attribute and **caret** attribute are right skewed, thus to ensure that our machine learning models are accurate and robut, data transformations will be conducted.


### 4.3 Data Transformation
```{r}
library(superml)
label <- LabelEncoder$new()
data$cut <- label$fit_transform(data$cut)
data$color <- label$fit_transform(data$color)
data$clarity <- label$fit_transform(data$clarity)
summary(data)
```

From the above summary, we can see that all categorical attributes (**cut**, **color**, **calarity**) were encoded into numerical values, this is crucial for the robustness of our machine learning models

Since the data attributes have different scales, standardization is required to ensure that our machine learning model is not facing overfitting/underfitting issues and biases
```{r}
model_data = scale(data)
summary(model_data)
```


Finally, the data will be split into 70% training set and 30% testing set
```{r}
set.seed(101) # Set Seed so that same sample can be reproduced 

# Now Selecting 70% of data as sample from total 'n' rows of the data  
sample <- sample.int(n = nrow(model_data), size = floor(.7*nrow(model_data)), replace = F)
train_data <- model_data[sample, ]
test_data  <- model_data[-sample, ]
```

<!-- STUDENT 3 AND 4  -->
<center> <h1><b>5.Machine Learning</b></h1> </center> 


<!-- STUDENT 1 -->
<center> <h1><b>6.Conclusion</b></h1> </center>
